\section{Open items}
\label{sec:open}

Two checklists: task-graph/algorithm correctness (Section~\ref{sec:todo-task})
and physical validation (Section~\ref{sec:todo-phys}). Move an item to
Section~\ref{sec:bugs} (if it was a bug) or
Section~\ref{sec:validation} (if it's a settled result) once resolved
-- do not just delete it here.

Several items below carry labels (\code{e6}--\code{e10}) inherited
from the pre-existing project memory
(\code{radiation\_feedback\_taskgraph\_plan.md}, sessions predating
this logbook) to avoid losing that history; new items from this
session are labelled \code{T}\emph{n} (task-graph) and
\code{V}\emph{n} (validation).

\subsection{Task graph / algorithm}
\label{sec:todo-task}

\begin{itemize}
  \item[$\square$] \textbf{T1} -- Confirm or rule out the
    \code{radiation\_level} orphaned-link risk
    (Section~\ref{sec:not-regular}) on the clump examples specifically.
    It requires two neighbouring regions' \code{hydro.super} to land
    at different tree depths, which uniform-density ICs never trigger
    but a $100\times$ density clump plausibly could. Rebuild with
    \code{--enable-debugging-checks} and run
    \code{StromgrenSphereClumpSmith2021} at a resolution fine enough
    to actually split the tree near the clump boundary; the
    \code{error()} at \code{cell.c:1387} will fire if it's hit.
  \item[$\boxtimes$] \textbf{e7 -- assessed, not an issue (2026-07-16)}
    -- Interior-recombination-accounting gap: once a particle is
    ionized it never reappears as a search candidate (it fails the
    ``cold'' filter in \code{feedback\_part\_can\_be\_ionized}) until
    its tag expires, so its ongoing maintenance recombination cost is
    never re-charged against a star's budget at any rebuild after the
    one that first tagged it. This matches FIRE's own implementation
    (not a deviation to fix), and is bounded by design here: this
    model's rebuild cadence
    (\code{HII\_region\_rebuild\_time\_Myr}) is set to track the
    recombination/cooling timescale, so interior gas has no time to
    meaningfully recombine away between rebuilds -- no example or test
    run so far has shown this gap actually matters in practice.
    \textbf{Possible refinement, not confirmed necessary:} add a small
    margin to \code{end\_time} (Section~\ref{sec:algorithm},
    Persistence) beyond the nominal
    \code{tag\_time + HII\_region\_rebuild\_time}, to cover cases
    where SWIFT's own timestepping forces a rebuild slightly earlier
    than the nominal cadence -- untested, low priority given the above.
  \item[$\square$] \textbf{e10} -- Verify the D-type (Spitzer) growth
    curve's \emph{shape}, not just its final-time value.
    \code{stromgren\_analytic\_check.py} (in the sibling
    \code{StromgrenSphere} example) already implements the analytic
    reference $R(t) = R_{\rm st}(1+7c_st/4R_{\rm st})^{4/7}$
    \citep{Spitzer1978} and reports a pass/fail relative error at final
    time for a single isolated star; it does not yet check the
    trajectory throughout. Run it on a single isolated star (no
    clump, no co-location) before drawing further conclusions from the
    8-run matrix's endpoint-only numbers.
  \item[$\boxtimes$] \textbf{T2 -- implemented (2026-07-16)} -- $\beta$
    (case-B recombination coefficient, Eq.~\ref{eq:recomb-cost}) was a
    fixed constant evaluated at the conventional $10^4\,$K
    (\code{phys\_const->const\_caseb\_recomb}), not recomputed at the
    gas's actual, metallicity-dependent applied temperature
    (Section~\ref{sec:metallicity}); Section~\ref{sec:beta-temperature}
    quantified the cost as a $\sim$75\% overestimate at this example's
    $Z=0$ default. Now replaced with \citet{HuiGnedin1997}'s
    $\alpha_B(T)$ fit (Eq.~\ref{eq:alphaB-huignedin}), evaluated at the
    particle's own applied ionized temperature -- see
    Section~\ref{sec:beta-temperature} for the implementation and
    Section~\ref{sec:validation} for its measured effect on the
    matrix.
  \item[$\square$] \textbf{T3} -- Decide the example's default for
    \code{HII\_deterministic\_boundary\_ionization}. Currently 1
    (deterministic), but Section~\ref{sec:validation}'s matrix shows
    deterministic mode over-ionizes -- mildly for the uniform scheme,
    severely for the angular scheme (drives it into the search-radius
    cap). Probabilistic (0) gives the cleanest, uncapped match to both
    papers. Not yet changed pending a decision on whether the example
    should default to the setting that best matches the papers, or to
    whichever setting is intended for production galaxy-scale runs
    (where the angular-cap interaction may not arise the same way).
  \item[$\square$] \textbf{T4 -- now confirmed to bind, not just rows
    2/6} -- \code{Stars:max\_HII\_search\_radius} (45\,pc) and the box
    (100\,pc, 50\,pc half-width) are too tight for the
    angular+deterministic combination, which reaches
    $\sim$37--45\,pc and gets artificially capped
    (Table~\ref{tab:matrix}, rows 2/6). Widen both together (e.g.\
    \citeauthor{Smith2021}'s own \code{rion\_max=100\,pc}) and rerun
    those two rows for an uncapped comparison. Since T2 (below), row 4
    -- previously comfortably uncapped -- now also hits this same
    ceiling (Section~\ref{sec:caseB-fix-validation}), so this item now
    blocks getting a converged T2 result too, not just rows 2/6.
  \item[$\square$] \textbf{T5} -- MPI is explicitly not supported yet
    for the radiation loop (\texttt{error("MPI is not yet
    implemented")} in both \code{task.c} and \code{cell\_unskip.c}
    for \texttt{nr\_nodes > 1}). Not a bug, just an explicit gap to be
    aware of before attempting a multi-node run with this feedback
    channel active.
\end{itemize}

\subsection{Physical validation}
\label{sec:todo-phys}

\begin{itemize}
  \item[$\square$] \textbf{V1} -- For all 8 matrix configurations
    (Table~\ref{tab:matrix}), not just the two plotted, check
    per-star, per-direction reach (clump-facing vs.\ away), not only
    the scalar max $r_{\rm HII}$ -- a single number can hide an
    asymmetric front the way it did before the radial-shell
    temperature check caught the clump's shallow surface erosion.
  \item[$\square$] \textbf{V2} -- Once T3 is decided, update
    \code{run.sh}/\code{params.yml} defaults and rerun the
    acceptance checklist (build, unit tests, both
    \code{StromgrenSphere} acceptance runs) with the new default.
  \item[$\square$] \textbf{V3} -- After T4, rerun the angular +
    deterministic + fast-rebuild configuration (row 2) with a widened
    cap/box to see its true, uncapped growth -- currently unknown
    whether it converges near the papers' range or keeps growing.
  \item[$\square$] \textbf{V4} -- Run e10's curve-shape check (above)
    before and after any future algorithm change to this model (e.g.\
    T2, or e7's optional \code{end\_time} margin if that's ever
    implemented), to confirm a change actually improves the trajectory
    shape and not just the endpoint by coincidence.
  \item[$\square$] \textbf{V5} -- Quantify whether the shallow clump
    surface erosion found this session (star-facing outer skin
    $\sim$28,000\,K to $\sim$3\,pc deep, interior still cold) grows
    toward \citeauthor{Smith2021}'s more substantial disruption given
    more time / the fixed rebuild cadence, or genuinely plateaus short
    of it -- not yet run long enough or compared quantitatively
    (their Fig.~2 doesn't give an exact eroded-mass number to target,
    only a visual).
  \item[$\square$] \textbf{V6} -- T2 is implemented
    (Section~\ref{sec:beta-temperature}); row 4 has been re-run with it
    (Section~\ref{sec:validation}). Re-run the \emph{remaining} 7 rows
    of Table~\ref{tab:matrix} to check whether T2 closes any of the
    residual gap to the papers' $\sim$20--25\,pc reference across the
    whole matrix, per the Section~\ref{sec:metallicity} prediction that
    it should only ever push our radius \emph{larger}.
  \item[$\square$] \textbf{V7} -- Pin down \citeauthor{Smith2021}'s
    actual box size for their idealized D-type test more precisely
    than the current eyeballed 60--80\,pc estimate (Section
    \ref{sec:validation}), if an exact box-size comparison becomes
    load-bearing for some future conclusion.
  \item[$\boxtimes$] \textbf{V8} -- Run the Smith clump example at the
    metallicity that makes $T_{\rm collisional}=10^4\,$K exactly
    ($Z/Z_\odot\approx0.231$, Eq.~\ref{eq:tcollisional}), isolating a
    validation point where the fixed-$\beta$ approximation
    (Section~\ref{sec:beta-temperature}) is exact by construction --
    cleanly separating ``is T2 the cause of any residual gap'' from
    other candidate causes, without needing to implement T2 first.
    Done 2026-07-16, see Section~\ref{sec:validation}.
  \item[$\square$] \textbf{V9} -- Root-cause the no-clump overshoot
    (Section~\ref{sec:no-clump}): removing the clump
    (\code{density\_factor=1}) from the row-4 configuration gives
    32.6--35.6\,pc at $t=8\,$Myr, \emph{larger} than the papers'
    $\sim$20--25\,pc no-clump reference, whereas the clumpy version of
    the same row lands right in that range. A clump can only remove
    budget from the diffuse medium, so its absence should not by
    itself produce an overshoot -- candidate explanations not yet
    checked: whether Hu/Smith report a spherically-averaged radius
    rather than our per-star, max-over-angular-pixel $r_{\rm HII}$;
    whether 4 literally co-located sources searching/consuming
    independently behaves differently from one combined-luminosity
    source; or a genuine resolution/box-size effect.
  \item[$\square$] \textbf{V10 -- priority follow-up to T2} -- Rerun
    row 4 with T2 applied at a widened
    \code{Stars:max\_HII\_search\_radius}/box (T4) to get an uncapped
    post-fix result: the current 42.9--44.9\,pc
    (Section~\ref{sec:caseB-fix-validation}) is very likely a lower
    bound, not a converged value (all four stars sit at or within
    2\,pc of the 45\,pc cap and none has flattened by $t=8\,$Myr). Only
    once uncapped can this be honestly compared to the papers'
    $\sim$20--25\,pc reference -- right now it is already well above
    that reference \emph{before} even reaching convergence, which on
    its own is worth flagging as a possible over-correction, but not
    yet confirmed as one.
\end{itemize}
